\jaxcosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-06-28 paper.tex
minor
- Abstract : first draft - cite PyMC as new PPL with JAX backend - New sections: CRediT authorship contribution statement & Declaration of Competing Interest following the model of https://www.sciencedirect.com/science/article/pii/S2213133722000373#d1e10839

@@

\begin{abstract}
With \jaxcosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial. \github
\end{abstract}

@@

Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process. \texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive. Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing. We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming language that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.

@@

@@

@@

%
%
\subsection{DES-Y1 Example}
%
%\Acknowledgements
\section*{Acknowledgements}
Some of the numerical experiments have been conducted at the IN2P3 Computing Center (CC-IN2P3 - Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique.

>>>>

\newcommand{\FL}[1]{{\color{cyan}FL: #1}}

@@

\begin{abstract}
The recent JAX library creates a new  ecosystem in which the developments of probabilistic programming software take benefit of the attractive speed up as batch acceleration, Just-In-Time compilation and optimization of code for different hardware (CPU, GPU, TPU). The JAX automatic differentiation mechanism makes more robust algorithms used in probability density sampling, eg. Hamiltonian like Monte Carlo as well as Variotional Inference. The current development of Bayesian Inference (with of without likelihood) with large data sets in Cosmology rises the opportunity to develop the \texttt{jax-cosmo} library \github. The current version allows to compute linear and non linear power spectra (\texttt{halofit}) with Eisenstein and Hu transfer function, as well as angular power spectra ($C_\ell$) in Limber approximation for galaxy and weak lensing probes. The comparison with the Core Cosmological Library is exposed. Then, we have conducted in the context of DES Y1 3x2pts analysis a series of cosmological parameters as well as nuisance parameters estimations using different MCMC methods using \numpyro\ as PPL framework. The mathematical background of the different methods is detailed. As result of the numerical experiments, We show a clear improvement using HMC compared to Metreoplosis-Hasting generation and that the Normalizing Flows using the Neutral Transport is very effective in the early phases of Bayesian model development. 
\end{abstract}

@@

Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process. \texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive. Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing. We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library,  \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming language (PPL) that is used in this paper. Notice also that other PPL softwares have switched to JAX backend as for instance the latest \texttt{PyMC} v4.0.1\footnote{\url{https://www.pymc.io/projects/docs/en/stable/installation.html}}.

@@

\FL{Add description of problem and small demos}

@@

\FL{mention something about scaling with parallel MCMC}

@@

%
%
\subsection{DES-Y1 Example}
%

%%%

{\color{red} the next sections are written along the model of \url{https://www.sciencedirect.com/science/article/pii/S2213133722000373} in Astronomy and Computing}
%%%
\section*{CRediT authorship contribution statement}
\textbf{F. Lanusse}, \textbf{J. Zuntz}...
\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.

\section*{Declaration of Competing Interest}
The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

%\Acknowledgements
\section*{Acknowledgements}
Some of the numerical experiments have been conducted at the IN2P3 Computing Center (CC-IN2P3 - Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique.
